\textbf{Measuring Depression and Anxiety with 4 items? Adaptation of the PHQ-4 to increase its Sensitivity to Subclinical Variability}

### 2025-06-09 paper/5_Collabra/manuscript.qmd
minor changes to manuscript

@@

The consistency of the anxiety ($Cronbach's~\alpha = 0.903$) and depression ($Cronbach's~\alpha = 0.841$) subscales is excellent (responses were treated as ordinal). The proportion of response types stratified by item (see @fig-one) shows that the new "Once or twice" option was the most prevalent response for all items (on average selected in 29.12\% of cases).

@@

Participants were randomly assigned to complete either the original or refined version of the PHQ-4, which included one additional response option ("Once or twice"), scored 0.5 to preserve the same scoring as the original version (creating more possible total scores - 0.5, 1.5, 2.5, etc.).

@@

**One of the potential limitation of our study includes the choice of the measures used as external "ground-truth" of depression and anxiety, namely the BDI-II and the STAI-5. For instance, there is ongoing debate about the STAI's  discriminant validity (which might extend to its short form used in the present study). A recent meta-analysis reported that trait anxiety scores were more strongly associated with depressive than anxiety disorders [@knowles2020specificity], suggesting that the scale may rather capture general negative affect than specifically anxiety. Regarding the BDI-II, existing evidence that suggests its relative lack of sensitivity for low-severity depression [@olino2012measuring] - comparable to similar instruments of its size (e.g., CES-D), and an improvement over the BDI-I [@wahl2014standardization] - put into question its choice for our goal of showing increased sub-clinical sensitivity. Thus, future studies should verify these findings with alternative measures of anxiety and depression and clinically assessed populations.**

>>>>

**Enabling a more precise assessment of low-severity alterations is important, as milder symptoms are significant predictors of future clinical disorders and are associated with present functional impairments and reduced quality of life [@judd1998major; @cuijpers2004subthreshold]. In addition, the growing reliance on large-scale, often online, psychological and epidemiological surveys, to monitor population-level mental health, evaluate interventions, or track responses to global stressors such as pandemics or geopolitical crises, demands tools that are both brief and sensitive to small but meaningful fluctuations.**

@@

The consistency of the anxiety ($Cronbach's~\alpha = 0.903$) and depression ($Cronbach's~\alpha = 0.841$) subscales is excellent (**responses were treated as ordinal**). The proportion of response types stratified by item (see @fig-one) shows that the new "Once or twice" option was the most prevalent response for all items (on average selected in 29.12\% of cases).

@@

Participants were randomly assigned to complete either the original or refined version of the PHQ-4, which included one additional response option ("Once or twice"). **To preserve comparability with the original PHQ-4 scoring system and avoid altering the scale’s total score range or established cut-off thresholds, we assigned the new response option a value of 0.5, placing it midway between "Not at all" (0) and "Several days" (1). While this scoring assumes equal spacing between response options - a common but imperfect convention in ordinal scales - it offers a pragmatic compromise between conceptual fidelity and applied utility.**

@@

**One of the potential limitations of our study includes the choice of the measures used as external "ground-truth" of depression and anxiety, namely the BDI-II and the STAI-5. For instance, there is ongoing debate about the STAI's  discriminant validity (which might extend to its short form used in the present study). A recent meta-analysis reported that trait anxiety scores were more strongly associated with depressive than anxiety disorders [@knowles2020specificity], suggesting that the scale may rather capture general negative affect than specifically anxiety. Regarding the BDI-II, existing evidence that suggests its relative lack of sensitivity for low-severity depression [@olino2012measuring] - comparable to similar instruments of its size (e.g., CES-D), and an improvement over the BDI-I [@wahl2014standardization] - put into question its choice for our goal of showing increased sub-clinical sensitivity. Thus, future studies should verify these findings with alternative measures of anxiety and depression and clinically assessed populations.**

### 2025-06-09 paper/5_Collabra/manuscript.tex
minor changes to manuscript

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{An Shu
Te}{0000-0002-9312-5552}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{S.H. Annabel
Chen}{0000-0002-1540-5516}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, colback=white, toprule=.15mm, opacityback=0, arc=.35mm, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, leftrule=.75mm]

@@

The consistency of the anxiety (\(Cronbach's~\alpha = 0.903\)) and
depression (\(Cronbach's~\alpha = 0.841\)) subscales is excellent
(responses were treated as ordinal). The proportion of response types
stratified by item (see Figure~\ref{fig-one}) shows that the new ``Once
or twice'' option was the most prevalent response for all items (on
average selected in 29.12\% of cases).

@@

\textbf{One of the potential limitation of our study includes the choice
of the measures used as external ``ground-truth'' of depression and
anxiety, namely the BDI-II and the STAI-5. For instance, there is
ongoing debate about the STAI's discriminant validity (which might
extend to its short form used in the present study). A recent
meta-analysis reported that trait anxiety scores were more strongly
associated with depressive than anxiety disorders
(\citeproc{ref-knowles2020specificity}{Knowles \& Olatunji, 2020}),
suggesting that the scale may rather capture general negative affect
than specifically anxiety. Regarding the BDI-II, existing evidence that
suggests its relative lack of sensitivity for low-severity depression
(\citeproc{ref-olino2012measuring}{Olino et al., 2012}) - comparable to
similar instruments of its size (e.g., CES-D), and an improvement over
the BDI-I (\citeproc{ref-wahl2014standardization}{Wahl et al., 2014}) -
put into question its choice for our goal of showing increased
sub-clinical sensitivity. Thus, future studies should verify these
findings with alternative measures of anxiety and depression and
clinically assessed populations.}

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{An Shu
Te}{0000-0002-9312-5552}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{S.H. Annabel
Chen}{0000-0002-1540-5516}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, left=2mm, opacityback=0, rightrule=.15mm, leftrule=.75mm, arc=.35mm, breakable, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, toprule=.15mm]

@@

The consistency of the anxiety (\(Cronbach's~\alpha = 0.903\)) and
depression (\(Cronbach's~\alpha = 0.841\)) subscales is excellent
(\textbf{responses were treated as ordinal}). The proportion of response
types stratified by item (see Figure~\ref{fig-one}) shows that the new
``Once or twice'' option was the most prevalent response for all items
(on average selected in 29.12\% of cases).

@@

\textbf{One of the potential limitations of our study includes the
choice of the measures used as external ``ground-truth'' of depression
and anxiety, namely the BDI-II and the STAI-5. For instance, there is
ongoing debate about the STAI's discriminant validity (which might
extend to its short form used in the present study). A recent
meta-analysis reported that trait anxiety scores were more strongly
associated with depressive than anxiety disorders
(\citeproc{ref-knowles2020specificity}{Knowles \& Olatunji, 2020}),
suggesting that the scale may rather capture general negative affect
than specifically anxiety. Regarding the BDI-II, existing evidence that
suggests its relative lack of sensitivity for low-severity depression
(\citeproc{ref-olino2012measuring}{Olino et al., 2012}) - comparable to
similar instruments of its size (e.g., CES-D), and an improvement over
the BDI-I (\citeproc{ref-wahl2014standardization}{Wahl et al., 2014}) -
put into question its choice for our goal of showing increased
sub-clinical sensitivity. Thus, future studies should verify these
findings with alternative measures of anxiety and depression and
clinically assessed populations.}
